Preschoolers form conventional pacts with each other to communicate about novel shapes

### 2025-01-16 write-ups/kid_tangrams_cogsci_2025/kid_tangrams_cogsci_2025.Rmd
start to address Alvin Mike comments

abstract: >
   Learning language requires learning not only the content of language, but also skills that allow us to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants jointly converge on mutually understandable names for initially novel objects. Some classical experiments with young children are interpreted as showing that 4-5-year old children are incapable of succeeding at iterated reference games. Here, we revisit the question of young children's referential communicative abilities using a simpler, child-friendly paradigm. Across 51 pairs of 4-5-year old children, we find that preschool-aged children were successful in an iterated reference game. Children were 85% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with a larger picture of the early emerging ability to pragmatically use language in communication.

@@

output: cogsci2024::cogsci_paper
header-includes: \cogscifinalcopy
---

@@

stats_text <- function(model, row, decimal = 2) {
  model <- model |>
    mutate(
      Estimate = round(Estimate, digits = decimal),
      Lower = round(lower, digits = decimal),
      Upper = round(upper, digits = decimal),
      `Credible Interval` = str_c("[", Lower, ", ", Upper, "]")
    ) |>
    select(Term, Estimate, `Credible Interval`)
  str_c(model[row, 2], "  ", model[row, 3])
}

@@

A classic study suggests that 4-5-year old preschoolers struggle with child-child referential communication [@glucksberg1966]. In their paradigm, one child was given a set of 6 blocks in a specific order. Their task was to describe the images on the block so their partner could pick out the corresponding block. As children described and selected blocks, they stacked them on pegs. While 4-5-year old children were successful on practice trials with familiar shapes and visual access to each other's blocks, children failed on critical trials where the blocks had drawings of abstract shapes and there was no visual access. Even after multiple rounds with the same images, children were not able to correctly order the blocks. <!--@glucksberg1966 attributed young children's failure to the children’s use of idiosyncratic referring expressions unique to their own experiences, making it difficult for the pair to converge on shared descriptions.-->
Similar experiments with older children indicated a gradual improvement through adolescence both at their initial accuracy and on their improvement over repetitions. Still, even the 9th grade sample was noticeably worse than the sample of adult college students [@glucksberg1967; @krauss1969]. That the task is difficult even among teenagers suggests that the complexity of the stacking paradigm and the large number of potential targets present significant task demands that might prevent accurate measurement of children's abilities.

@@

While children are sensitive to others’ perspectives, they still seem to struggle with appropriately tailoring the specificity of their utterances to the visual context, resulting in utterances that are under-informative [@matthews2012; @leung2024].  When children must describe one of two very similar images, 4 and 5-year olds struggle to mention the relevant features, although they do better when playing in an interactive game than when not [@grigoroglou2019]. By 5-years old, children are sensitive to utterances that are over- or under-informative, asking for clarification on some under-informative utterances and taking longer to make selections on over-informative utterances [@morisseau2013].

@@

Since @glucksberg1966, few studies have revisited the question of whether children can successfully communicate in an iterated reference game. Recent work shows that 8-10-year olds exhibit adult-like patterns of increasing accuracy, increasing speed, and shorter descriptions across repetitions, but children’s accuracy was still far below adult performance  and highly variable between dyads [@branigan2016]. 4-6-year olds have also been shown to rely on conventionalized gestures with their partner in an iterated reference game where children could only use gestures to communicate [@bohn2019].

A recent study with younger children examined 4-8 year olds' abilities to play an iterated reference game with a parent using a simple, child-friendly tablet-based task. In this task, even 4-year olds had an initial accuracy above 80%, which rose to above 90% in later repetitions [@leung2024]. Together, these findings provide evidence of young children’s ability to successfully communicate about novel referents under the right conditions.

Given the task demands in @glucksberg1966 and work showing that children can succeed in a simplified, less demanding paradigm, we revisit the question of child-child referential communication. In the present study, we re-examine young children's ability to establish effective referring expressions with each other in an iterated reference game using a simplified tablet-based paradigm. Across 51 pairs of 4-5-year old children, we found that preschool-aged children were successful in an iterated reference game, suggesting that children's capacity to construct effective referring expressions in novel contexts emerges earlier than once claimed.

# Experiment 1 Methods

@@

## Participants
4 and 5-year old children were recruited from a university preschool during the school day. Children played with another child from the same class. Experiment 1 was conducted between June and August 2023. Pairs of children were included in analyses if they completed at least 8 of the 12 critical trials. We had 19 complete games and 1 included incomplete game. Of the 40 children, 21 were girls, and the median age was 57 months, with a range of 48-70 months.

@@

## Materials
For the target stimuli, we used four of the ten tangram images from @leung2024, chosen based on visual dissimilarity (Figure \ref{game}B). We coded the matching game using Empirica and hosted it on a lab server [@almaatouq2020]. We then accessed the game on the web on tablets that were locked in a kiosk mode so children could not navigate away from the game.

## Procedure
Once a pair of children agreed to play the game, a research assistant took them to a quiet testing room. Children were introduced to a stuffed animal "Smurfy" who wanted to play a matching game. Children sat across a table from each other, each with a tablet in front of them (Figure \ref{game}A). On each trial, one child was the "teller" and saw a black box around one of two images on their screen and was asked to "say what they saw" in the black box. The "guesser" saw the same two images in a randomized order and tried to select the described image to help Smurfy make a match. When the guesser selected an image, both children received feedback in form of a smiley or frowny face and an audible sound. After each trial, children switched roles. Children passed Smurfy back and forth to help them keep track of whether they were the "guesser" or "teller" on a given trial.

@@

## Data processing

@@

# Experiment 1 Results

## Accuracy

```{r fig-accuracy, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Children's accuracy at selecting the correct target over time. Experiment 1 had 3 blocks (12 total critical trials) and experiment 2 had 4 blocks (16 critical trials). Error bars are bootstrapped 95% CIs with a linear trend line overlaid. \\label{acc}" }
clean_2 |>
  filter(type != "practice") |>
  mutate(trial = trial - 3, expt = "Expt 2") |>
  bind_rows(clean_1 |> filter(type != "practice") |> mutate(trial = trial - 1, expt = "Expt 1")) |>
  ggplot(aes(x = trial, y = correct)) +
  stat_summary(aes(fill = type), fun.data = "mean_cl_boot", geom = "bar") +
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange", color = "black") +
  scale_fill_manual(values = time_palette) +
  # geom_bar(position = position_dodge(.9), color = "black", fill = "#19ADDE", width = .7) +
  geom_smooth(method = "lm", color = "black") +
  # geom_text(aes(label = percent), hjust = 1.6, color = "white", size = 3) +
  xlab("Trial") +
  ylab("Accuracy") +
  scale_x_continuous(breaks = seq(1, 16, 4), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(0, 1, .2), expand = c(0, 0)) +
  geom_hline(yintercept = .5) +
  facet_grid(expt ~ .) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

Our primary measure of interest was whether children could accurately communicate the intended target. <!-- as prior work is often interpreted as indicating the children at kindergarten age cannot communicate about abstract shapes successfully [@glucksberg1966]. -->  As shown in Figure \ref{acc}, children were above chance in their selections. To test for changes in accuracy over time, we fit a mixed effects model of accuracy (`r acc_form |> form()`). Children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_1,1)`) and their accuracy slightly increased over the game (OR of one trial later: `r stats_text(acc_mod_1, 2)`).

## Speed

@@

<!--[less important, could drop] --> As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects model: `r speed_form |> form()`. The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)` seconds / trial ).

## Description length

@@

In iterated reference games with adults, description lengths usually shorten over repeated references [@clark1986; @hawkins2020b; @boyce2024]. We were curious if children's descriptions would display the same trend, so we ran a Bayesian mixed effects model of how long of a description the "teller" produced: `r description_form |> form()`. On the first critical trial, descriptions averaged `r stats_text(description_mod_1, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_1, 2)` words per trial, Figure \ref{length}).

How long of a description is initially warranted depends on how easily describable the target is, as well as how large and close the contrast set is. In this case, the low number of options and relatively easy to describe shapes may have meant long initial descriptions were less necessary. The adult controls in @leung2024 use shorter initial descriptions than adults in studies with larger arrays of harder to distinguish images. However, young children may also choose to provide shorter descriptions than adults would, and young children may accept shorter descriptions as adequate whereas adults may ask for more information to increase their confidence level before making a guess. <!-- some of this is kinda discussion-y, possibly move -->

## Convergence

@@

Different children in the same game produced more similar descriptions than children in different games, although the descriptions were less similar than descriptions from the same child in different rounds (Figure \ref{sim}). We modeled this as `r sim_across_between_form |> form()`. Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_1_results |> stats_text(3,3)`). The big differences in descriptions between games compared to within games is a measure of partner sensitivity.

# Experiment 1 Discussion

@@

# Experiment 2 Methods

## Participants
Experiment 2 was run between March and August of 2024, at the same university preschool as experiment 1. No children participated in both experiments. 30 pairs of children completed all 16 critical trials, and 1 pair of children completed between 8 and 16 critical trials. Our target age range was 4 and 5 year olds, but one almost 4-year-old was unintentionally included. Of the 62 children, 30 were girls, and the children had a median age of 56 months, and a range of 45-69 months.

## Materials
The same 4 critical images were used as in experiment 1, although this time, children saw these images 4 times. In response to some children struggling with the abrupt switch from familiar to non-nameable shapes, we introduced more practice trials for experiment 2. We used a total of 4 practice trials to provide a gradient from familiar shapes to less recognizable, blockier shapes (Figure \ref{game}D).

## Procedure
The procedure was much the same as Experiment 1(Figure \ref{game}D). We added an initial "bubble popping" exercise to give children practice with how to tap the tablet appropriately (this was an issue for some children in Experiment 1). The experimental script was fully written out and memorized by experimenters so children all received the same instructions. We wrote up contingency statements that the experimenter could use to prompt children who were not giving descriptions or making selections. Experimenters helped with game mechanics such as whose turn it was to tell and who should press the screen to move the game along, but avoided contributing or repeating any content about the images or the descriptions.

@@

```
## Data processing
Data was processed in the same way as experiment 1. After excluding trials where children did not give a description or where the experimenter echoed a child's description, we had `r clean_2 |> filter(!is.na(repNum)) |> nrow()` trials total.

# Experiment 2 Results

We report the same set of analyses and model results as in Experiment 1 as well as additional analyses of how the semantic content of children's descriptions changes over time.

## Accuracy

@@

As in Experiment 1, children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_2,1)`) and accuracy slightly increased over the course of the game (OR of one trial later: `r stats_text(acc_mod_2, 2)`, Figure \ref{acc}). This confirms that children are able to communicate with each other about these abstract shapes.

## Speed

@@

In Experiment 2, the first critical trial averaged `r stats_text(speed_mod_2, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_2, 2)`seconds / trial). Children were initially faster in Experiment 2 than Experiment 1, possibly due to the increased number of practice trials and pre-training on how to press the screens.

## Description length

@@

Children used a wide range of descriptions successfully, and some example descriptions from Experiment 2 are shown in Table \ref{example}. Often children would use the same description as their partner, although there were some transitions such as from "a person holding a flute" to "a person holding a trumpet".

## Convergence

@@

```{r sim2, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Semantic similarity between earlier blocks (1-3) with last block (4) for descriptions to the same image within the same group. SImilarity measured as cosine similarity between S-BERT embeddings of referential descriptions. Heavy dots are means with bootstrapped 95% CIs; light dots are individual values. \\label{tolast}" }
sims_2 |>
  filter(game1 == game2) |>
  mutate(later = ifelse(block1 > block2, block1, block2), earlier = ifelse(block1 > block2, block2, block1)) |>
  filter(later == 4) |>
  mutate(same_speaker = ifelse(speaker1 == speaker2, "same speaker", "different speaker")) |>
  ggplot(aes(as.character(earlier), sim, color = target1)) +
  geom_point(alpha = .1, position = position_dodge(width = .2)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  facet_wrap(~same_speaker) +
  scale_color_solarized() +
  labs(x = "Earlier block to last block", y = "cosine similarity", color = "target") +
  theme(legend.position = "bottom")
```

@@

To look for convergence to shared descriptions within games, we compared the utterances from the first three blocks to the descriptions in the last block: `r sim_to_last_form |> form()`. Over the first three blocks, descriptions became increasingly similar to the last block description (`r sim_to_last_2_results |> stats_text(2,3)`). Descriptions were more similar to the last block if they came from the same child who gave the description in the last block (`r sim_to_last_2_results |> stats_text(3,3)`). 
<!--Another way to look for convergence is to look at the semantic distance between utterances in adjacent blocks: `r sim_to_next_form |> form()`. --> Although over time descriptions do get more similar to the last block utterance, the semantic distance between adjacent block utterances was relatively constant: `r sim_to_next_2_results |> stats_text(2,3)`.

As each partnership converged to their shared nicknames, partnerships often diverged from one another as groups focused on distinct aspects of the image. We tested whether descriptions in different games to the same target diverged over time: `r sim_across_form |> form()`. As the games progressed, descriptions from different games became slightly further apart in semantic space (`r sim_across_2_results |> stats_text(2,3)`).

# Additional results

## Interaction structure
Pairs of children varied significantly in how well they could scaffold their own interaction versus needed reminders from experimenters about the structure of the game. In one game, the experimenter provided no input after the practice trials because the children rapidly took turns producing referring expressions and selecting the targets. Their advantage could partially be due to being best friends with each other; all children were paired with another child from the same classroom who they were willing to play with, but friendship levels varied.

@@

## Meta-analytic models

<!--Combining across experiments

<!-- Might want to split out and incorporate in along with expt 2 analyses? -->

@@

As the two experiments were similar to one another, we re-ran models pooling across the two experiments. 
Pooling the two experiments, children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_meta,1)`) and accuracy slightly increased over the course of the game (OR of one trial later: `r stats_text(acc_mod_meta, 2)`. 
The average length of descriptions on the first trial was `r stats_text(description_mod_meta, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_meta, 2)` words per trial).

@@

@@

In the broader picture of language acquisition, there's debate over whether children learn the facts of their language first, followed much later, by pragmatic and communicative competency (CITATIONS), or whether sensitivity to communicative intent is an early emerging skill that develops in parallel with linguistic knowledge and may bootstrap language learning (CITATIONS). The findings in the current work are most consistent with a gradual development of children's communicative and linguistic skills, where the skills emerge early and then are refined over time, as children's cognitive capacities increase. At 4-5-years old, children are able to use their linguistic skills communicatively, even as their utterances are not fully adult-like in form.

>>>>

abstract: >
   Learning language requires learning not only the content of language, but also skills that allow us to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants jointly converge on mutually understandable names for initially novel objects. Some classical experiments with young children are interpreted as showing that 4-5-year old children are incapable of succeeding at iterated reference games. Here, we revisit the question of young children's referential communicative abilities using a simpler, child-friendly paradigm. Across 51 pairs of 4-5-year old children, we find that preschool-aged children were successful in establishing reference. Children were 85% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with the view that children show early pragmatic competence in supportive contexts.

@@

output: cogsci2024::cogsci_paper
header-includes: \raggedbottom
#header-includes: \cogscifinalcopy
---

@@

stats_text <- function(model, row, decimal = 2) {
  model <- model |>
    mutate(
      Estimate = round(Estimate, digits = decimal) |> formatC(format='f', digits=decimal ),
      Lower = round(lower, digits = decimal) |> formatC(format='f', digits=decimal ),
      Upper = round(upper, digits = decimal) |> formatC(format='f', digits=decimal ),
      `Credible Interval` = str_c("[", Lower, ", ", Upper, "]")
    ) |>
    select(Term, Estimate, `Credible Interval`)
  str_c(model[row, 2], "  ", model[row, 3])
}

@@

A classic study suggests that 4-5-year old preschoolers struggle with child-child referential communication [@glucksberg1966]. In their paradigm, one child was given a set of 6 blocks in a specific order. Their task was to describe the images on the block so their partner could pick out the corresponding block. As children described and selected blocks, they stacked them on pegs. While 4-5-year old children were successful on practice trials with familiar shapes and visual access to each other's blocks, children failed on critical trials where the blocks had drawings of abstract shapes and there was no visual access. Even after multiple rounds with the same images, children were not able to correctly order the blocks. <!--@glucksberg1966 attributed young children's failure to the children’s use of idiosyncratic referring expressions unique to their own experiences, making it difficult for the pair to converge on shared descriptions.-->
Similar experiments with older children indicated a gradual improvement through adolescence both for initial accuracy and for the increase in accuracy across repetitions. Still, even the 9th grade sample was noticeably worse than the sample of adult college students [@glucksberg1967; @krauss1969]. That the task was difficult even among teenagers suggests that the complexity of the stacking paradigm and the large number of potential targets present significant task demands that might prevent accurate measurement of children's abilities.

@@

While children are sensitive to others’ perspectives, they still seem to struggle with appropriately tailoring the specificity of their utterances to the visual context, often resulting in utterances that are under-informative [@matthews2012; @leung2024].  When children must describe one of two very similar images, 4 and 5-year olds struggle to mention the relevant features, although they do better when playing in an interactive game than when not [@grigoroglou2019]. By 5-years old, children are sensitive to utterances that are over- or under-informative, asking for clarification on some under-informative utterances and taking longer to make selections on over-informative utterances [@morisseau2013].

@@

Since @glucksberg1966, few studies have revisited the question of whether children can successfully communicate in an iterated reference game. Recent work shows that 8-10-year olds exhibit adult-like patterns of increasing accuracy, increasing speed, and shorter descriptions across repetitions, but children’s accuracy was still far below adult performance  and highly variable between dyads [@branigan2016]. 4-6-year olds have also been shown to rely on conventionalized gestures with their partner in an iterated reference game where children could only use gestures to communicate [@bohn2019]. Recent work has established that young children can succeed at iterated reference games when playing with their parents [@leung2024]. 4-8 year olds' played an iterated reference game with a parent using a simple, child-friendly tablet-based task. In this task, even 4-year olds had an initial accuracy above 80%, which rose to above 90% in later repetitions [@leung2024]. Together, these findings provide evidence of young children’s ability to successfully communicate about novel referents under the right conditions.

Given the task demands in @glucksberg1966 and work showing that children can succeed in a less demanding paradigm, we revisit the question of child-child referential communication. In the present study, we re-examine young children's ability to establish effective referring expressions with each other in an iterated reference game using a simplified tablet-based paradigm. Across 51 pairs of 4-5-year old children, we found that preschool-aged children were successful in an iterated reference game, suggesting that children's capacity to construct effective referring expressions in novel contexts emerges earlier than once claimed.

# Experiment 1

## Methods

@@

### Participants
4 and 5-year old children were recruited from a university preschool during the school day. Children played with another child from the same class. Experiment 1 was conducted between June and August 2023. Pairs of children were included in analyses if they completed at least 8 of the 12 critical trials. We had 19 complete games and 1 included incomplete game. Of the 40 children, 21 were girls, and the median age was 57 months, with a range of 48-70 months.

@@

### Materials
For the target stimuli, we used four of the ten tangram images from @leung2024, chosen based on visual dissimilarity (Figure \ref{game}B). We coded the matching game using Empirica and hosted it on a lab server [@almaatouq2020]. We then accessed the game on the web on tablets that were locked in a kiosk mode so children could not navigate away from the game.

### Procedure
Once a pair of children agreed to play the game, a research assistant took them to a quiet testing room. Children were introduced to a stuffed animal "Smurfy" who wanted to play a matching game. Children sat across a table from each other, each with a tablet in front of them (Figure \ref{game}A). On each trial, one child was the "teller" and saw a black box around one of two images on their screen and was asked to "say what they saw" in the black box. The "guesser" saw the same two images in a randomized order and tried to select the described image to help Smurfy make a match. When the guesser selected an image, both children received feedback in form of a smiley or frowny face and an audible sound. After each trial, children switched roles. Children passed Smurfy back and forth to help them keep track of whether they were the "guesser" or "teller" on a given trial.

@@

### Data processing

@@

## Results

### Accuracy

```{r fig-accuracy, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Children's accuracy at selecting the correct target over time. Experiment 1 had 3 blocks (12 total critical trials) and experiment 2 had 4 blocks (16 critical trials). Error bars are bootstrapped 95% CIs with a linear trend line overlaid. \\label{acc}" }
clean_2 |>
  filter(type != "practice") |>
  mutate(trial = trial - 3, expt = "Expt 2") |>
  bind_rows(clean_1 |> filter(type != "practice") |> mutate(trial = trial - 1, expt = "Expt 1")) |>
  ggplot(aes(x = trial, y = correct)) +
    geom_smooth(method = "lm", color = "black") +
  stat_summary(aes(color = type), fun.data = "mean_cl_boot") +
  #stat_summary(fun.data = "mean_cl_boot", geom = "linerange", color = "black") +
  scale_color_manual(values = time_palette) +
  # geom_bar(position = position_dodge(.9), color = "black", fill = "#19ADDE", width = .7) +
  # geom_text(aes(label = percent), hjust = 1.6, color = "white", size = 3) +
  xlab("Trial") +
  ylab("Accuracy") +
  geom_vline(xintercept = 4.5, lty="dashed")+
    geom_vline(xintercept = 8.5, lty="dashed")+
    geom_vline(xintercept = 12.5, lty="dashed")+
  scale_x_continuous(breaks = seq(1, 16, 4), expand = c(0.02, 0.02)) +
  scale_y_continuous(breaks = seq(0, 1, .2), expand = c(0, 0.04)) +
  coord_cartesian(ylim = c(0.3,1))+
  geom_hline(yintercept = .5) +
  facet_grid(expt ~ .) +
  theme(
    legend.position = "none",
    axis.ticks.x=element_blank(),
    axis.text.x=element_blank(),
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

Our primary measure of interest was whether children could accurately communicate the intended target. <!-- as prior work is often interpreted as indicating the children at kindergarten age cannot communicate about abstract shapes successfully [@glucksberg1966]. -->  As shown in Figure \ref{acc}, children were above chance in their selections. To test for changes in accuracy over time, we fit a Bayesian mixed effects logistic regression predicting accuracy (`r acc_form |> form()`). Children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_1,1)`) and their accuracy slightly increased over the game (OR of one trial later: `r stats_text(acc_mod_1, 2)`).

### Speed

@@

<!--[less important, could drop] --> As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects linear regression predicting the time to selection in seconds: `r speed_form |> form()`. The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)` seconds / trial).

### Description length

@@

In iterated reference games with adults, description lengths usually shorten over repeated references [@clark1986; @hawkins2020b; @boyce2024]. We were curious if children's descriptions would display the same trend, so we ran a Bayesian mixed effects linear regression predicting the number of words in the description the "teller" produced: `r description_form |> form()`. On the first critical trial, descriptions averaged `r stats_text(description_mod_1, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_1, 2)` words per trial, Figure \ref{length}).

### Convergence

@@

Different children in the same game produced more similar descriptions than children in different games, although the descriptions were less similar than descriptions from the same child in different rounds (Figure \ref{sim}). We modeled this as a Bayesian mixed effects linear regression predicting similarity `r sim_across_between_form |> form()`. Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_1_results |> stats_text(3,3)`). The big differences in descriptions between games compared to within games is a measure of partner sensitivity.

## Discussion

@@

@@

# Experiment 2

## Methods

### Participants
Experiment 2 was run between March and August of 2024, at the same university preschool as experiment 1. No children participated in both experiments. 30 pairs of children completed all 16 critical trials, and 1 pair of children completed between 8 and 16 critical trials. Our target age range was 4 and 5 year olds, but one almost 4-year-old was unintentionally included. Of the 62 children, 30 were girls, and the children had a median age of 56 months, and a range of 45-69 months.

### Materials
The same 4 critical images were used as in experiment 1, although this time, children saw these images 4 times. In response to some children struggling with the abrupt switch from familiar to non-nameable shapes, we introduced more practice trials for experiment 2. We used a total of 4 practice trials to provide a gradient from familiar shapes to less recognizable, blockier shapes (Figure \ref{game}D).

### Procedure
The procedure was much the same as Experiment 1(Figure \ref{game}D). We added an initial "bubble popping" exercise to give children practice with how to tap the tablet appropriately (this was an issue for some children in Experiment 1). The experimental script was fully written out and memorized by experimenters so children all received the same instructions. We wrote up contingency statements that the experimenter could use to prompt children who were not giving descriptions or making selections. Experimenters helped with game mechanics such as whose turn it was to tell and who should press the screen to move the game along, but avoided contributing or repeating any content about the images or the descriptions.

### Data processing
Data was processed in the same way as experiment 1. After excluding trials where children did not give a description or where the experimenter echoed a child's description, we had `r clean_2 |> filter(!is.na(repNum)) |> nrow()` trials total.

## Results

We report the same set of analyses and model results as in Experiment 1 as well as additional analyses of how the semantic content of children's descriptions changes over time. Children used a wide range of descriptions successfully, and some example descriptions from Experiment 2 are shown in Table \ref{example}.

### Accuracy

@@

In Experiment 2, children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_2,1)`) and accuracy was relatively stable over the course of the game (OR of one trial later: `r stats_text(acc_mod_2, 2)`, Figure \ref{acc}). This confirms that children are able to communicate with each other about these abstract shapes.

### Speed

@@

In Experiment 2, the first critical trial averaged `r stats_text(speed_mod_2, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_2, 2)` seconds / trial). Children were initially faster in Experiment 2 than Experiment 1, possibly due to the increased number of practice trials and pre-training on how to press the screens.

### Description length

@@

### Convergence

@@

@@

To look for convergence to shared descriptions within games, we compared the utterances from the first three blocks to the descriptions in the last block using a Bayesian mixed effects linear regression predicting similarity: `r sim_to_last_form |> form()`. Over the first three blocks, descriptions became increasingly similar to the last block description (`r sim_to_last_2_results |> stats_text(2,3)`). Descriptions were more similar to the last block if they came from the same child who gave the description in the last block (`r sim_to_last_2_results |> stats_text(3,3)`). 
<!--Another way to look for convergence is to look at the semantic distance between utterances in adjacent blocks: `r sim_to_next_form |> form()`. --> Although over time descriptions do get more similar to the last block utterance, the semantic distance between adjacent block utterances was relatively constant: `r sim_to_next_2_results |> stats_text(2,3)`.

As each partnership converged to their shared nicknames, partnerships often diverged from one another as groups focused on distinct aspects of the image. We tested whether descriptions in different games to the same target diverged over time using a Bayesian mixed effects linear regression: `r sim_across_form |> form()`. As the games progressed, descriptions from different games became slightly further apart in semantic space (`r sim_across_2_results |> stats_text(2,3)`).

```{r sim2, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Semantic similarity between earlier blocks (1-3) with last block (4) for descriptions to the same image within the same group. SImilarity measured as cosine similarity between S-BERT embeddings of referential descriptions. Heavy dots are means with bootstrapped 95% CIs; light dots are individual values. \\label{tolast}" }
sims_2 |>
  filter(game1 == game2) |>
  mutate(later = ifelse(block1 > block2, block1, block2), earlier = ifelse(block1 > block2, block2, block1)) |>
  filter(later == 4) |>
  mutate(same_speaker = ifelse(speaker1 == speaker2, "same speaker", "different speaker")) |>
  ggplot(aes(as.character(earlier), sim, color = target1)) +
  geom_point(alpha = .1, position = position_dodge(width = .2)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  facet_wrap(~same_speaker) +
  scale_color_solarized() +
  labs(x = "Earlier block to last block", y = "cosine similarity", color = "target") +
  theme(legend.position = "bottom")
```
# Additional results

## Qualitative examples of interaction structure
Pairs of children varied significantly in how well they could scaffold their own interaction versus needed reminders from experimenters about the structure of the game. In one game, the experimenter provided no input after the practice trials because the children rapidly took turns producing referring expressions and selecting the targets. Their advantage could partially be due to being best friends with each other; all children were paired with another child from the same classroom who they were willing to play with, but friendship levels varied.

@@

## Pooling across experiments

@@

As the two experiments were similar to one another, we re-ran models pooling the data across the two experiments. 
Pooling the two experiments, children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_meta,1)`) and accuracy slightly increased over the course of the game (OR of one trial later: `r stats_text(acc_mod_meta, 2)`). 
The average length of descriptions on the first trial was `r stats_text(description_mod_meta, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_meta, 2)` words per trial).

@@

It is unclear to what extent the uniformly short descriptions are a product of the simplified task or children's behavioral differences from adults. The appropriate length for an initial description depends on how easily describable the target is, as well as how large and close the contrast set is. In this case, the low number of options and relatively easy to describe shapes may have meant long initial descriptions were less necessary. The adult controls in @leung2024 used shorter initial descriptions than adults in studies with larger arrays of harder to distinguish images. However, young children may also choose to provide shorter descriptions than adults would, and young children may accept shorter descriptions as adequate whereas adults may ask for more information to increase their confidence level before making a guess. Especially in light of other work suggesting that conceptual pact formation and reduction in utterance length sometimes decouple in adults [@boyce2024], further empirical work on the factors driving verbosity in reference games is warranted.

@@

In the broader picture of language acquisition, there's debate over whether children learn the facts of their language first, followed much later, by pragmatic and communicative competency (CITATIONS), or whether sensitivity to communicative intent is an early emerging skill that develops in parallel with linguistic knowledge and may bootstrap language learning (CITATIONS). The findings in the current work are most consistent with a gradual development of children's communicative and linguistic skills, where the skills emerge early and then are refined over time, as children's cognitive capacities increase. At 4–5 years old, even though their utterances are not fully adult-like in form, children are nonetheless able to use their linguistic skills communicatively.
